Towards Practical First-Order Model Counting

### 2024-02-01 doc/paper.tex
Minor changes

\usepackage{listings}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{microtype}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{tikz}

@@

% TODO: mention/describe the arithmetic simplification code

% TODO: (in Fig. 1) Crane2 doesn't use Crane, it uses itself.

% TODO: describe which compilation rules are currently greedy or non-greedy.

@@

\paragraph{Proof of Correctness}
\begin{enumerate}
  \item Define what it means for a formula to be \emph{consistent}.
        \begin{itemize}
          \item The formula (including the mapping of constants to their
                domains) does not mention a domain together with its subdomain
                (i.e., all mentioned domains are pairwise disjoint).
                Specifically, we can't have a subdomain together with its parent
                domain or a subdomain together with a constant that belongs to
                its parent domain.
          \item The target of \texttt{Ref} must also be a consistent formula.
          \item The initial formula is consistent by definition.
        \end{itemize}
  \item Aim to show that Crane works correctly on all consistent formulas.
  \item Prove that, whenever some compilation rule makes the formula $\phi$
        inconsistent, greedy compilation rules will make $\phi$ consistent again
        before $\phi$ encounters non-greedy compilation rules that depend on
        $\phi$ being consistent.
        \begin{itemize}
          \item In particular, prove that shattering and unit propagation will
                always eliminate the parent domain of the domains introduced by
                atom counting.
          \item Constraint removal only applies when it can eliminate the parent
                domain entirely.
          \item Prove this for a particular assumption about which compilation
                rules are set to be greedy. For the greedy algorithm, although
                all rules are applied in a greedy manner, we still assume that
                greedy rules will be applied before non-greedy rules.
        \end{itemize}
  \end{enumerate}

@@

\paragraph{Contributions.}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item The experiments are (going to be) more comprehensive than any other
        experimental study performed on (W)FOMC algorithms (TODO: references to
        my previous paper and some of the \textsc{ForcLift} and
        \textsc{FastWFOMC} papers that contain experimental results).
\end{itemize}

@@

\paragraph{Notation.}
We write $\expr{}$ for an arbitrary algebraic expression. For any signature or
clause $C$, argument or variable $x$, and number or constant $t$, we shall write
$C[t / x]$ for the result of substituting $t$ for all occurrences of $x$ in $C$.
% TODO: check if I'm still using this notation in all these cases
% TODO: explain how this works for function calls too

\begin{definition}
  A \emph{function call} is a term of the form
  $f(x_{1} - c_{1}, \dots, x_{n} - c_{n})$ (written $f(\mathbf{x} - \mathbf{c})$
  for short), where $f$ is an $n$-ary function, each $x_{i}$ is a variable, and
  each $c_{i}$ is a non-negative constant.
\end{definition}
% TODO: do I ever use this term for the case when some of the variables don't
% exist?

@@

\begin{definition}
  An \emph{equation} is always of the form $f(\mathbf{x}) = \expr{}$, where
  $f(\mathbf{x})$ is a signature, and $\expr{}$ is an algebraic expression.
  Henceforth, we call $f(\mathbf{x})$ and $\expr{}$ the left-hand side (LHS) and
  the right-hand side (RHS) of the equation, respectively.
\end{definition}
% TODO: give some examples of formulas that: 1) introduce new variables, 2) call
% the function itself, 3) call some other function, 4) is more nested than just
% a sum.

@@

% TODO: what notation am I using for FOL? Colon after every quantifier?

% TODO: (Similar to CNF, maybe QBF). Incomplete. Need (W)FOMC semantics. What is
% an interpretation? What is a model?

A \emph{formula} is a set of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\forall x_{2} \in \Delta_{2}\dots\forall x_{n} \in \Delta_{n} \phi(x_{1}, x_{2}, \dots, x_{n})$,
where $\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). A \emph{literal} is either an atom or
its negation. An \emph{atom} is of the form $P(t_{1}, \dots, t_{m})$, where $P$
is a predicate, and $t_{1}, \dots, t_{m}$ are terms. A \emph{term} is either a
variable or a constant.
% TODO: explain set-based notation for a clause (iterating over all literals,
% maybe I should have a Lits() operator? No!!!)

@@

\begin{algorithm}[t]
  \caption{Finding the definitions of the base cases}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{a set $E$ of equations that define $B$}
  $(E, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$E$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(E', \textunderscore, \textunderscore) \gets \Crane{$\psi$}$\;
    $E \gets E \cup E'$\;
  }
\end{algorithm}

@@

% TODO: extend it to work with base cases that are themselves recursive

% TODO: what does it mean for a set of equations to define a set of base cases?

% TODO: for base cases, we will also use the f(\mathbf{x}) notation

@@

\begin{itemize}
  \item \cref{line:functioncall} iterates over all function calls in the
        algebraic expression.
  \item note that on \cref{line:signature} it is the signature and not the
        original function call.
\end{itemize}

% TODO: remove the 'dependencies' data structure from the description

@@

\begin{example}
  TODO: add an example where \cref{alg1} runs on either
  \begin{align*}
    f_{0}(m, n) &= f_{1}(m-1, n) + f_{2}(m, n-1)\\
    f_{1}(m, n) &= f_{1}(m-1, n-1) \times f_{2}(m-2, n-1)\\
    f_{2}(m, n) &= 2 \times f_{1}(m-3, n-1)
  \end{align*}
  or
  \begin{align*}
    f(m, n) &= g(m-1, n) + f(m-2, n-1)\\
    g(m, n) &= f(m-1, n-2) + g(m-1, n-1).
  \end{align*}

  \begin{itemize}
    \item \cref{line:lim}: the limit is 2 for the arg $(x-3)$.
    \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
          $n = 0$, add $f(y, z)[y/0] = f(0, z)$ to $B$.
  \end{itemize}
\end{example}

TODO: I changed the algorithm

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;
      }% TODO: find the source code and check: 1) what we're doing with unit clauses, 2) whether we just randomly pick any literal to negate
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}
% TODO: it would be easy to extend the second part to an arbitrary constant

@@

% TODO: The report has some (long) examples of formulas being transformed into
% programs perhaps suitable for supplementary material.

>>>>

\usepackage{listings}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{microtype}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage[backgroundcolor=lightgray]{todonotes}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item mention/describe the arithmetic simplification code
    \item (in Fig. 1) Crane2 doesn't use Crane, it uses itself
    \item describe which compilation rules are currently greedy or non-greedy
  \end{itemize}
}

@@

\paragraph{Proof of correctness.}
\begin{enumerate}
  \item Define what it means for a formula to be \emph{consistent}.
        \begin{itemize}
          \item The formula (including the mapping of constants to their
                domains) does not mention a domain together with its subdomain
                (i.e., all mentioned domains are pairwise disjoint).
                Specifically, we can't have a subdomain together with its parent
                domain or a subdomain together with a constant that belongs to
                its parent domain.
          \item The target of \texttt{Ref} must also be a consistent formula.
          \item The initial formula is consistent by definition.
        \end{itemize}
  \item Aim to show that Crane works correctly on all consistent formulas.
  \item Prove that, whenever some compilation rule makes the formula $\phi$
        inconsistent, greedy compilation rules will make $\phi$ consistent again
        before $\phi$ encounters non-greedy compilation rules that depend on
        $\phi$ being consistent.
        \begin{itemize}
          \item In particular, prove that shattering and unit propagation will
                always eliminate the parent domain of the domains introduced by
                atom counting.
          \item Constraint removal only applies when it can eliminate the parent
                domain entirely.
          \item Prove this for a particular assumption about which compilation
                rules are set to be greedy. For the greedy algorithm, although
                all rules are applied in a greedy manner, we still assume that
                greedy rules will be applied before non-greedy rules.
        \end{itemize}
  \end{enumerate}

@@

\paragraph{Contributions.}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item The experiments are (going to be) more comprehensive than any other
        experimental study performed on (W)FOMC algorithms.
\end{itemize}

\todo[inline]{references to my previous paper and some of the \textsc{ForcLift}
  and \textsc{FastWFOMC} papers that contain experimental results}

@@

\paragraph{Notation.}
We write $\expr{}$ for an arbitrary algebraic expression. For any signature or
clause $C$, argument or variable $x$, and number or constant $t$, we shall write
$C[t / x]$ for the result of substituting $t$ for all occurrences of $x$ in $C$.

\todo[inline,caption={}]{
  \begin{itemize}
    \item check if I'm still using this notation in all these cases
    \item explain how this works for function calls too
  \end{itemize}
}

\begin{definition}
  A \emph{function call} is a term of the form
  $f(x_{1} - c_{1}, \dots, x_{n} - c_{n})$ (written $f(\mathbf{x} - \mathbf{c})$
  for short), where $f$ is an $n$-ary function, each $x_{i}$ is a variable, and
  each $c_{i}$ is a non-negative constant.
\end{definition}
\todo[inline]{do I ever use this term for the case when some of the variables
  don't exist?}

@@

\begin{definition}
  An \emph{equation} is always of the form $f(\mathbf{x}) = \expr{}$, where
  $f(\mathbf{x})$ is a signature, and $\expr{}$ is an algebraic expression.
  Henceforth, we call $f(\mathbf{x})$ and $\expr{}$ the left-hand side (LHS) and
  the right-hand side (RHS) of the equation, respectively.
\end{definition}
\todo[inline,caption={}]{
  give some examples of formulas that:
  \begin{itemize}
    \item introduce new variables,
    \item call the function itself,
    \item call some other function, and
    \item is more nested than just a sum.
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item What notation am I using for FOL? Colon after every quantifier?
    \item (Similar to CNF, maybe QBF). Incomplete. Need (W)FOMC semantics. What
          is an interpretation? What is a model?
  \end{itemize}
}

A \emph{formula} is a set of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\forall x_{2} \in \Delta_{2}\dots\forall x_{n} \in \Delta_{n} \phi(x_{1}, x_{2}, \dots, x_{n})$,
where $\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). A \emph{literal} is either an atom or
its negation. An \emph{atom} is of the form $P(t_{1}, \dots, t_{m})$, where $P$
is a predicate, and $t_{1}, \dots, t_{m}$ are terms. A \emph{term} is either a
variable or a constant.

\todo[inline]{explain set-based notation for a clause (iterating over all
  literals)}

@@

\begin{algorithm}[t]
  \caption{Finding the definitions of the base cases}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{a set $E$ of equations that define $B$}
  $(E, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$E$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(E', \text{\textunderscore}, \text{\textunderscore}) \gets \Crane{$\psi$}$\;
    $E \gets E \cup E'$\;
  }
\end{algorithm}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item extend it to work with base cases that are themselves recursive
    \item what does it mean for a set of equations to define a set of base cases?
    \item for base cases, we will also use the $f(\mathbf{x})$ notation
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item NOTE: \cref{line:functioncall} iterates over all function calls in the
          algebraic expression
    \item remove the `dependencies' data structure from the description
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  add an example where \cref{alg1} runs on either
  \begin{align*}
    f_{0}(m, n) &= f_{1}(m-1, n) + f_{2}(m, n-1)\\
    f_{1}(m, n) &= f_{1}(m-1, n-1) \times f_{2}(m-2, n-1)\\
    f_{2}(m, n) &= 2 \times f_{1}(m-3, n-1)
  \end{align*}
  or
  \begin{align*}
    f(m, n) &= g(m-1, n) + f(m-2, n-1)\\
    g(m, n) &= f(m-1, n-2) + g(m-1, n-1).
  \end{align*}
}

\begin{itemize}
  \item \cref{line:lim}: the limit is 2 for the arg $(x-3)$.
  \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
        $n = 0$, add $f(y, z)[y/0] = f(0, z)$ to $B$.
\end{itemize}

\todo[inline]{I changed the algorithm. Update the description.}

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;
      }% TODO: find the source code and check: 1) what we're doing with unit clauses, 2) whether we just randomly pick any literal to negate
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

\todo[inline]{it would be easy to extend the second part to an arbitrary constant}

@@

\todo[inline]{the report has some (long) examples of formulas being transformed
  into programs perhaps suitable for supplementary material}
